On Optimality of Finite Element Integration

### 2015-11-05 TOMS_Paper/assembly.tex
Minor fixes

\subsubsection{Observations and underlying assumption}
\label{sec:se-assumptions}
What we are advocating is the fundamental fact that there never are two ore more conflicting factorizations among which to choose for minimizing the operation count. The idea is that sub-expressions must be placed within the loops that they logically belong to (and not in a deeper level), and this can only be accomplished by systematic factorization and code motion.

There only are two pathological circumstances in which this approach might not lead to optimality: 
\begin{itemize}
\item As already explained in Section~\ref{sec:se-opln}, having both test and trial functions in the same function space exposes two alternative transformations.
\item The loops degenerate (i.e., their trip count is 1). This makes the factorizations described in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln} sub-optimal since code motion is now irrelevant. For example, the reader can verify that expanding the expression $(a f_i + b g_i) (c f_i + d g_i)$ and factorizing $L_i$-dependent symbols is optimal if and only if the size of $L_i$ is strictly greater than 1 (the sub-expression was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient).
\end{itemize}

\subsubsection{Algorithm}
Algorithm~\ref{algo:opt} describes how we manipulate the input loop nest and expression for performing sharing elimination. Sections~\ref{sec:se-opln} and~\ref{sec:se-rln} can be seen as a constructive proof of optimality under the (reasonable) assumptions in Section~\ref{sec:se-assumptions}.

@@

We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms. The choice of what we should factorize first impacts the resulting operation count. It depends on two factors: the number of unique terms in each loop, $n$, and the trip counts, $tc$. It is fairly simple to see (and easily demonstrable by contradiction) that by starting from the loop such that $n \cdot tc$ is smallest leads to the maximal operation count reduction.

Assume, without loss of generality, that we start with $L_j$. Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression. The other sub-expression, independent of $L_j$ ($L_k$), is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting.

For each of the expressions lifted at the level of $L_i$, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions are hoisted. Every time a sub-expression is hoisted, we search and replace all common sub-expressions. This last step is probably practically useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.
\end{Algo}

@@

In the previous section, we provided an insight into the potentially negative effects of code motion. In this section, we expand on this matter. The following two observations, in particular, lead to the definition of \textit{memory constraints}.

@@

\subsection{Trading optimality for analysis cost}
Code generation time is fundamental for productivity. In this section, we discuss how to reduce it by relaxing optimality.
\label{sec:obsheu}
\begin{itemize}
\item \textit{Sharing elimination.} In Section~\ref{sec:sharing-elimination}, we explained that very small loop trip counts may render certain factorizations sub-optimal. To reach optimality, Algorithm~\ref{algo:opt} needs to backtrack if it detects an increase in operation count after applying sharing elimination at a given loop nest level. Since such increases are, in practice, rare and negligible, the algorithm can be relieved by avoiding these checks.
\item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible combinations in which pre-evaluation candidates can be scheduled to determine the optimal configuration honoring memory constraints. Despite the fact that we have never observed any tangible increase in analysis cost, one could expect slow-downs in complex forms with many monomials. This matter is yet to be properly investigated. However, a possibility is to avoid or constrain the search; in the worst case scenario, this would be equal to reaching local optimums.
\item \textit{Construction of $\theta$.} Instead of determining exact values for $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by (partly) executing Algorithms~\ref{algo:opt} and~\ref{algo:pre-evaluation}, approximations can be found by exploiting loop linearity. The number of unique terms in a multilinear loop nest can be used to calculate a lower bound on the innermost loop operation count (where, for higher order methods, the bulk of the computation is performed). For example, in Figure~\ref{code:multi_loopnest_a}, only one term depends on $L_j$ ($b_j$), so at most one multiplication is expected in the innermost loop once sharing is eliminated. An approximation for $\theta_{se}$ is then $\theta_{se} = n_{iters} \cdot n_{unique}$, where $n_{iters}$ is the total number of loop iterations, and $n_{unique}$ is the smallest number of unique terms depending on one the linear loops.
\end{itemize}

>>>>

\subsubsection{Observations and underlying assumptions}
\label{sec:se-assumptions}
We observe that there almost never are two ore more conflicting factorizations among which to choose for minimizing the operation count. The idea is that, by choosing a factorization that places a sub-expression within the loop that it logically belongs to, a reduction in operation count is generally obtained. There only are two pathological circumstances in which such a reduction may not be obtained: 
\begin{itemize}
\item As explained in Section~\ref{sec:se-opln}, having both test and trial functions in the same function space exposes sub-expressions over different loops, which introduces a choice between two alternative transformations.
\item The trip count of a loop is extremely small (typically it needs be 1 for the following to happen). This decreases or nullifies the impact of code motion -- exposed through factorization, as described in Sections~\ref{sec:se-opln} and~\ref{sec:se-rln}. For instance, the reader can verify that expanding the expression $(a f_i + b g_i) (c f_i + d g_i)$ within a loop $L_i$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient} and factorizing $L_i$-dependent symbols leads to optimality if and only if the size of $L_i$ is strictly greater than 1.
\end{itemize}

\subsubsection{Algorithm}
Algorithm~\ref{algo:opt} details the steps for sharing elimination assuming as input an abstract syntax tree representation of the loop nest. Neglecting the pathological scenarios described in the previous section, the algorithm performs all necessary transformations and checks to reach optimality. A constructive proof of optimality can be derived taking into account both the algorithm itself and the observations in Sections~\ref{sec:se-opln},~\ref{sec:se-rln} and~\ref{sec:se-assumptions}.

@@

We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms. The choice of starting with $L_j$ or $L_k$ depends on two parameters: the number of unique terms, $n_{L}$, and the trip count, $tc_{L}$, of a loop $L$. It is fairly simple to see (and easily demonstrable by contradiction) that taking first $L$ such that $n_L \cdot tc_L$ is smallest leads to the largest operation count reduction.

Assume, without loss of generality, that we start with $L_j$. Thanks to linearity, each factored product only has one symbol depending on $L_j$, and this symbol is now unique in the expression tree. The other sub-expression, independent of $L_j$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting. We repeat the same process with $L_k$.

For each of the expressions lifted at the level of $L_i$, we factorize $L_i$-dependent terms such that $L_e$-dependent sub-expressions can be hoisted. Every time a sub-expression is hoisted, we search and replace all common sub-expressions. This last step is probably useless for the actual operation count, since we expect any general-purpose compiler to be able to perform aggressive common sub-expression elimination\footnote{By inspection of assembly code, we have verified that this is actually the case with the Intel compiler v15.2.}, but it helps making the generated code much more readable.
\end{Algo}

@@

In the previous section, we provided an insight into the potentially negative effects of code motion. In this section, we expand on this matter. The two following observations, in particular, lead to the definition of \textit{memory constraints}.

@@

\subsection{Trading optimality for analysis cost}
Code generation time is fundamental for productivity. In this section, we discuss how to reduce it by relaxing optimality.
\label{sec:obsheu}
\begin{itemize}
\item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible combinations in which pre-evaluation candidates can be scheduled to determine the optimal configuration honoring memory constraints. Despite the fact that we have never observed any tangible increase in analysis cost, one could expect slow-downs in complex forms with many monomials. This matter is yet to be properly investigated. However, a possibility is to avoid or constrain the search; in the worst case scenario, this would be equal to reaching local optimums.
\item \textit{Construction of $\theta$.} Instead of determining exact values for $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by (partly) executing Algorithms~\ref{algo:opt} and~\ref{algo:pre-evaluation}, approximations can be found by exploiting loop linearity. The number of unique terms in a multilinear loop nest can be used to calculate a lower bound on the innermost loop operation count (where, for higher order methods, the bulk of the computation is performed). For example, in Figure~\ref{code:multi_loopnest_a}, only one term depends on $L_j$ ($b_j$), so at most one multiplication is expected in the innermost loop once sharing is eliminated. An approximation for $\theta_{se}$ is then $\theta_{se} = n_{iters} \cdot n_{unique}$, where $n_{iters}$ is the total number of loop iterations, and $n_{unique}$ is the smallest number of unique terms depending on one the linear loops.
\end{itemize}
